Modelling BCG vaccination in the UK

### 2019-08-01 rmds/04-tb-epi-england.Rmd
minor revisions for TB epi from Ellen

---
output:
  pdf_document: default
---
# The epidemiology of tuberculosis, and the role of BCG vaccination, in England {#data}

@@

Whilst the characteristics of tuberculosis (TB) in England have been reported elsewhere,[@PHE2017; @PHE2018] and key risk factors such as non-UK birth status have been identified,@French2007 little attention has been given to the role of BCG vaccination. In particular, there is little information available regarding the demographics of vaccinated versus unvaccinated cases and the impact of BCG vaccination on TB outcomes in England has not been explored. There has also only been limited reporting of the age distribution, and trends over time, in incidence rates stratified by UK birth status.

In this chapter I explore the epidemiology of TB in England using routine data-sets, with a particular focus on the impact of missing data, the mechanisms underlying that missing date, seasonal trends, the role of age, UK birth status and BCG status. I have also estimated incidence rates, stratified by UK birth status and age, which I then used to identify trends in TB incidence over time. Finally I report TB outcomes in England using case rates, again stratified by BCG status and UK birth status. These data will then be used throughout this thesis to explore the impact of BCG vaccination on TB outcomes (Chapter \@ref(beneficial-bcg-out)), to estimate the direct impact of the 2005 change in BCG vaccination policy (Chapter \@ref(direct-eff)), to parameterise a dynamic TB transmission model (Chapter \@ref(model-development)) and to fit the same model to data (Chapter \@ref(model-fitting)).

@@

### Enhanced tuberculosis surveillance system {#ets-deep-dive}

@@

The enhanced tuberculosis surveillance system (ETS) is maintained by Public Health England (PHE) and collects demographic, clinical, and microbiological data on all notified cases in England. Notification is required by law, with health service providers having to inform PHE of all confirmed TB cases.[@PHE2017] Data collection began in 2000 and was expanded, with additional variables, with the launch of a web based system in 2008.[@Kriujshaar2007] It is updated annually with de-notifications, late notifications and other updates. A descriptive analysis of TB epidemiology in England is published each year, which reports on data collection, cleaning, and trends in TB incidence at both a national, and sub-national level.[@PHE2017]

@@

Data on all notifications from the ETS system from 2000 to 2015 were obtained from PHE. Data fields included: notification date, age, PHE centre, occupation, ethnic group, UK birth status, years since entry to the UK, date of symptom onset, date of presentation, date of diagnosis, date of treatment start, date of treatment end, date of death, Pulmonary TB status, culture status, sputum smear status, drug resistance, BCG vaccination status, year of vaccination, outcome at 12 months, overall outcome, and cause of death. Invalid entries were replaced with missing values unless otherwise noted, with character variables stored as factors using their most common entry as the baseline. Notifications from Scotland, Northern Ireland and Wales were dropped from the data-set. Several variables were created, or modified, for use in further analysis, Table \@ref(tab:created-var) summarises these variables. The code used for data cleaning is available as an R package ([https://www.samabbott.co.uk/tbinenglanddataclean/reference/clean_munge_ets_2016.html](https://www.samabbott.co.uk/tbinenglanddataclean/reference/clean_munge_ets_2016.html)).

```{r created-var}
created_var <- tibble(
  `Created/modified variable` = c('Years since BCG', 
                                  'Age at BCG', 
                                  'Successful treatment',
                                  'Mortality', 
                                  'TB mortality',
                                  'Death due to TB'),
  Description = c('Derived using year of vaccination and year of notification. Categorised into $\\leq 10$ and 11+ due to the evidence of waning protection for the BCG vaccine.[@Abubakar2013]', 
                  'Derived using year of vaccination and age at vaccination. Categorised into < 1, 1 to x < 12, 12 to x < 16 and $\\leq 16$ to capture historic vaccination policy.[@PublicHealthEngland2011a]', 
                  'For cases that had a recorded date of starting treatment, with their outcome recorded at the latest available follow up. Those that completed treatment are defined as successfully treated; treatment failure is defined as those that stopped treatment, were lost to follow up, those that died during follow up from TB, those that died during follow up were TB contributed to their death, and those who were still on treatment. Those that were not evaluated were treated as missing.', 
                  'Assessed via follow up at 12 and 24 months; mortality is defined as cases with an overall outcome of died, and survival is defined as those that completed treatment, were still on treatment, and stopped treatment. Those that were lost to follow up, or not evaluated were treated as missing', 
                  'For cases with an overall outcome of died, and whose cause of death was known to be TB or to be related to TB. Those that were known to have not died, or who were known to have died from a cause other than from TB were defined to have not died from TB.',
                  'Death due to TB is defined as those that died directly from TB, or where TB had contributed to their death with death not due to TB being cases that died from any other cause. Conditioned on all-cause mortality, for cases with a known cause of death.')

@@

Missing data can take several forms, data that is missing completely at random (MCAR), data that is missing at random (MAR) and data that is missing not at random (MNAR).@Sterne2009a Data that is MAR is missing with a mechanism that is conditional on observed variables, whilst MNAR is missing with a mechanism that is conditional on variables that are not observed. Data that is MAR, and MNAR may lead to biases when analysing the data, however it is not possible to deduce from the observed data what the mechanism driving missing data is. Therefore, it is necessary to account for these potential biases during the analysis stage. This is possible using a variety of methods such as scenario analysis accounting for the 'best' and 'worst' case scenario's, and multiple imputation of missing data using additional variables in the data set to inform the imputation model.@Sterne2009a

As the ETS is aggregated across England, from a variety of sources, some level of missing data is inevitable. This takes two forms; under-reporting of notified cases, of which there is some evidence in the literature,[@Pillaye2003] and data missing for a notified case. The former is particularly problematic as apart from using comparative studies the structure of those that are not notified is unknown. For variables that are missing data within the data-set it is possible to calculate the proportion of missing data (Figure \@ref(fig:plot-missing-struct), Table \@ref(tab:missing-var-tabs)) but care must be taken to account for nested variables such as date of death and year of BCG vaccination. This can be done by assuming the value for the top level variable when it is known that the variable is not truly missing. An example of this is using overall outcome for data of death when notifications are known to have not died.  Doing this shows high completeness for common demographic variables such as sex, age, ethnic group and UK birth status. More problematically, BCG status and year of BCG status have a high percentage missing, even after accounting for the introduction of national collection of these variables in 2008. Socio-economic status (as national quintiles) was not collected until 2010 but after this point is highly complete.  Comparing pre 2009 and post 2008 in Table \@ref(tab:missing-var-tabs) (and by inspecting Figure \@ref(fig:plot-missing-struct)) there are also issues of changing completeness over time,[@PHE2016; @PHE2017] if this is not accounted for than it may lead to spurious trends. Figure \@ref(fig:plot-missing-struct) also indicates that there are multiple groups of variables that share a common pattern of missing data.

```{r plot-missing-struct, fig.scap = "Summary plot of missing data in the extract of the ETS data used in this thesis.", fig.cap = "Summary plot of missing data in the extract of the ETS data used in this thesis. Due to the large size of the dataset, the data has been sub-sampled with only 20\\% of the data shown in this figure. Notifications have been ordered by date of notification from left to right. The following subset of variables are shown; year (year), sex (sex), age (age), PHE Centre (phec), Occupation (occat), Ethnic group (ethgrp), UK birth status (ukborn), Time since entry (timesinceent), date of symptom onset (symptonset), date of diagnosis (datediag), started treatment (startedtreat), date of starting treatment (starttreatdate), treatment end date (txenddate), pulmonary or extra-pulmonary TB (pulmextrapulm), culture (culture), sputum smear status (sputsmear), drug resistance (anyres), previous diagnosis (prevdiag), BCG status(bcgvacc), Year of BCG vaccination (bcgvaccyr), overall outcome (overalloutcome), cause of death (tomdeathrelate), socio-economic status quintiles (natquintile), and date of death (dateofdeath). Nested variables have been accounted for (i.e data of death has had an entry added for cases that are known to have not died), so that true missingness for all variables is estimated.", out.width = "60%", out.extra = ""}
knitr::include_graphics("chapters/tb-epi-england/figures/plot-missing-struct.png")
```

@@

kable(miss_ets,
      caption.short = "Breakdown of missing data from the ETS prior to the web based system (pre 2009) and post (post 2008) by variable, ordered by the  percentage missing for a subset of variables.",
      caption = "Breakdown of missing data from the ETS prior to the web based system (pre 2009) and post (post 2008) by variable, ordered by the  percentage missing for a subset of variables. The following subset of variables are shown; year (year), sex (sex), age (age), PHE Centre (phec), Occupation (occat), Ethnic group (ethgrp), UK birth status (ukborn), Time since entry (timesinceent), date of symptom onset (symptonset), date of diagnosis (datediag), started treatment (startedtreat), date of starting treatment (starttreatdate), treatment end date (txenddate), pulmonary or extra-pulmonary TB (pulmextrapulm), culture (culture), sputum smear status (sputsmear), drug resistance (anyres), previous diagnosis (prevdiag), BCG status(bcgvacc), Year of BCG vaccination (bcgvaccyr), overall outcome (overalloutcome), cause of death (tomdeathrelate), socio-economic status quintiles (natquintile), and date of death (dateofdeath). Nested variables have been accounted for (i.e data of death has had an entry added for cases that are known to have not died), so that true missingness for all variables is estimated.",
      booktabs = TRUE) %>%
  add_header_above(c(" ", "Pre 2009" = 2, "Post 2008" = 2)) %>% 
  kable_styling(latex_options = "HOLD_position") 
```

For nested variables, with rare outcomes, assuming the top level variable value can mask the underlying amount of missing data. An alternative approach is to filter the data for the top level variable required for the nested variable to be defined and to then compute the proportion of these notifications that are missing the data for the outcome of interest. For the date of starting treatment this approach leads to an estimate of `r missing_stats[["date_treat"]]` being missing, which is more complete than previously estimated. For cases that are known to have completed treatment `r missing_stats[["date_treat_end"]]` are missing a date for the end of treatment. In notifications that are known to have died, `r missing_stats[["date_death"]]` were missing the date of death and `r missing_stats[["cause_death"]]` were missing the cause of death. In any analysis were these variables are used the missing data for these variables will need to be carefully adjusted for. In particular, if cause of death is used it must be clearly stated that it is highly missing and results based on this variable should be properly caveated.

@@

As previously discussed, missing data may be MAR or MNAR, which may introduce biases into any analyses based on these data. This is of particular importance for variables that have high levels of missingness, as any introduced bias is likely to have a greater impact on the overall results, and for variables that are used extensively in analyses later in this thesis. Unfortunately MNAR data cannot be detected, so bias from this source cannot be discounted. However, it is possible to detect MAR mechanisms from observed variables that would not necessarily be included in a model used for analysis. For this reason, in the following section, I explore variables associated with data being missing for several key variables including: BCG status, year of BCG vaccination, date of death, cause of death, date of symptom onset, date of diagnosis, date of starting treatment and date of ending treatment. All of these variables were shown to have high levels of missing data in the previous section and they will all be used extensively throughout this thesis.

In order to explore the drivers of missing data I have reformulated the problem as a logistic regression for each variable of interest, with the outcome being data completeness (complete/missing). This allows variables that are hypothesised to be related to missing data to be adjusted for and their independent impact on data completeness to be measured. Unlike classic approaches to missing data, such as multiple imputation by chained regression (MICE),[@Groothuis-oudshoorn] this is not an imputation but insteaqd allows associations between observed varibales and missing outcome data to be estimated. The details of the approach are discussed below.

@@

In order to reformulate missing data as a logistic regression the following steps must be taken,

@@

1. Fit a logistic regression model; with the temporary data status variable as the outcome, adjusting for the hypothesised drivers of missingness.

1. Exponentiate the returned coefficients, and confidence intervals, so that they represent Odds Ratios (ORs).

@@

For all outcomes considered I adjusted for the same set of demographic variables that where both highly complete and also plausibly linked to missingness for all outcomes considered. They are; year, sex, age (grouped as 0-14 year old's, 15-65 year old's and 65+), ethnic group, UK birth status and socio-economic status (national quintiles). For socio-economic group 1 indicates the most deprived quintile. Complete case analysis has been used, with the dataset limited to notifications from 2010 and on-wards as socio-economic status was not collected prior to this.

@@

It is clear that BCG status is missing with a MAR mechanism for the variables considered (Table \@ref(tab:bcgvacc-miss)). BCG data being missing is strongly associated with year of notification, sex age, ethnic group, and socio-economic status. It appears that after adjusting for other variables data completeness increased from 2010 until 2012 but has since showed no clear trend. Men appear to be more likely than women to have a missing BCG status, with the non-UK born also being more likely than the UK born to be missing BCG status. The proportion of those missing BCG status increases with age, with those aged 65+ being over 4 times more likely to be missing BCG status than those aged 0-14 years old. There is also evidence to suggest that notifications in the lowest socio-economic group are more likely to have a missing BCG status but there was no clear evidence of a trend across socio-economic quintiles. The White ethnic group was more likely to have a missing BCG status than any other ethnic group.

```{r bcgvacc-miss}
kable(model_miss_results[["bcgvacc"]],
      caption.short = "Results from a logistic regression model with data completeness (Complete/Missing) for BCG vaccination as an outcome, adjusted for; year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles).",
      caption = "Results from a logistic regression model with data completeness (Complete/Missing) for BCG vaccination as an outcome, adjusted for; year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles). For socio-economic group 1 indicates the most deprived quintile. Notifications from 2010 onwards were included as socio-economic status was not collected before this. Complete case analysis weas used. Odds ratios shown are adjusted for all explanatory variables. The model indicates that BCG status is missing at random for the variables considered.", booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(font_size = 8, 
                latex_options = c("repeat_header")) %>% 
  column_spec(c(1), width = "1.5cm") %>% 
  column_spec(c(4), width = "2cm") %>% 
  column_spec(c(6, 7), width = "1.5cm") %>% 
  kable_styling(latex_options = "hold_position")
```

@@

```{r bcgvaccyr-miss}
kable(model_miss_results[["bvcvaccyr"]], format = "latex",
      caption.short = "Results from a logistic regression model with data completeness (Complete/Missing) for year of BCG vaccination as an outcome, adjusted for; year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles).",
      caption = "Results from a logistic regression model with data completeness (Complete/Missing) for year of BCG vaccination as an outcome, adjusted for; year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles). For socio-economic group 1 indicates the most deprived quintile. Notifications from 2010 onwards were included as socio-economic status was not collected before this. Complete case analysis weas used. Odds ratios shown are adjusted for all explanatory variables. The model indicates that year of BCG vaccination is missing at random for the variables considered.", booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(font_size = 8, latex_options = c("repeat_header")) %>% 
  column_spec(c(1), width = "1.5cm") %>% 
  column_spec(c(4), width = "2cm") %>% 
  column_spec(c(6, 7), width = "1.5cm")
```

@@

```{r dateofdeath-miss}
kable(model_miss_results[["dateofdeath"]],
      caption.short = "Results from a logistic regression model with data completeness (Complete/Missing) for date of death as an outcome, adjusted for; year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles).",
      caption = "Results from a logistic regression model with data completeness (Complete/Missing) for date of death as an outcome, adjusted for; year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles). For socio-economic group 1 indicates the most deprived quintile. Notifications from 2010 onwards were included as socio-economic status was not collected before this. Complete case analysis weas used. Odds ratios shown are adjusted for all explanatory variables. The model indicates that there is some evidence that date of death is missing at random for ethnic group, with weaker evidence for all other variables.", booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(font_size = 8, latex_options = c("repeat_header")) %>% 
  column_spec(c(1), width = "1.5cm") %>% 
  column_spec(c(4), width = "2cm") %>% 
  column_spec(c(6, 7), width = "1.5cm") %>% 
  kable_styling(latex_options = "hold_position")
```

@@

```{r tomdeathrealat-miss}
kable(model_miss_results[["tomdeathrealat"]],
      caption.short = "Results from a logistic regression model with data completeness (Complete/Missing) for cause of death as an outcome, adjusted for; year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles).",
      caption = "Results from a logistic regression model with data completeness (Complete/Missing) for cause of death as an outcome, adjusted for; year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles). For socio-economic group 1 indicates the most deprived quintile. Notifications from 2010 onwards were included as socio-economic status was not collected before this. Complete case analysis weas used. Odds ratios shown are adjusted for all explanatory variables. The model indicates that cause of death is missing at random for ethnic group and UK birth status, with little evidence for any other variables", booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(font_size = 8, latex_options = c("repeat_header")) %>% 
  column_spec(c(1), width = "1.5cm") %>% 
  column_spec(c(4), width = "2cm") %>% 
  column_spec(c(6, 7), width = "1.5cm") %>% 
  kable_styling(latex_options = "hold_position")
```

>>>>

---
output:
  pdf_document: default
---
# The epidemiology of tuberculosis and the role of BCG vaccination in England {#data}

@@

Although the characteristics of tuberculosis (TB) in England have been reported elsewhere,[@PHE2017; @PHE2018] and key risk factors such as non-UK birth status have been identified,@French2007 little attention has been given to the role of BCG vaccination. In particular, there is little information available regarding the demographics of vaccinated versus unvaccinated cases and the impact of BCG vaccination on TB outcomes in England has not been explored. There has also only been limited reporting of the age distribution, and trends over time, in incidence rates stratified by UK birth status.

In this chapter I explore the epidemiology of TB in England using routine data-sets, with a particular focus on the impact of missing data, the mechanisms underlying that missing data, seasonal trends, the role of age, UK birth status and BCG status. I have also estimated incidence rates, stratified by UK birth status and age, which I then used to identify trends in TB incidence over time. Finally I report TB outcomes in England using case rates, again stratified by BCG status and UK birth status. These data will then be used throughout this thesis to explore the impact of BCG vaccination on TB outcomes (Chapter \@ref(beneficial-bcg-out)), to estimate the direct impact of the 2005 change in BCG vaccination policy (Chapter \@ref(direct-eff)), to parameterise a dynamic TB transmission model (Chapter \@ref(model-development)) and to fit the same model to data (Chapter \@ref(model-fitting)).

@@

### Enhanced tuberculosis surveillance (ETS) system {#ets-deep-dive}

@@

The ETS is a database that collects demographic, clinical, and microbiological data on all notified TB cases in England and is maintained by Public Health England (PHE). Notification is required by law, with health service providers having to inform PHE of all confirmed TB cases.[@PHE2017] Data collection began in 2000 and was expanded, with additional variables, with the launch of a web based system in 2008.[@Kriujshaar2007] It is updated annually with de-notifications, late notifications and other updates. A descriptive analysis of TB epidemiology in England is published each year, which reports on data collection, cleaning, and trends in TB incidence at both a national, and sub-national level.[@PHE2017]

@@

Data on all notifications (114,820 notifications) from the ETS system from 2000 to 2015 were obtained from PHE via an application to the TB monitoring team. Data fields included: notification date, age, PHE centre, occupation, ethnic group, UK birth status, years since entry to the UK, date of symptom onset, date of presentation, date of diagnosis, date of treatment start, date of treatment end, date of death, Pulmonary TB status, culture status, sputum smear status, drug resistance, BCG vaccination status, year of vaccination, outcome at 12 months, overall outcome, and cause of death. Notifications were assessed for identifiability and the data release was conditional on the raw data not being shared further. Invalid entries were replaced with missing values unless otherwise noted, with character variables stored as factors using their most common entry as the baseline. Notifications from Scotland, Northern Ireland and Wales were dropped from the data-set. Several variables were created, or modified, for use in further analysis, Table \@ref(tab:created-var) summarises these variables. The code used for data cleaning is available as an R package^[Data cleaning code: https://www.samabbott.co.uk/tbinenglanddataclean/reference/clean_munge_ets_2016.html].

```{r created-var}
created_var <- tibble(
  `Created/modified variable` = c('Years since BCG', 
                                  'Age at BCG', 
                                  'Successful treatment',
                                  'Mortality', 
                                  'TB mortality',
                                  'Death due to TB'),
  Description = c('Derived using year of vaccination and year of notification. Categorised into $\\leq 10$ and 11+ due to the evidence of waning protection for the BCG vaccine.[@Abubakar2013]', 
                  'Derived using year of vaccination and age at vaccination. Categorised into < 1, 1 to x < 12, 12 to x < 16 and $\\leq 16$ to capture historic vaccination policy.[@PublicHealthEngland2011a]', 
                  'For cases that had a recorded date of starting treatment, with their outcome recorded at the latest available follow up. Those that completed treatment are defined as successfully treated: treatment failure is defined as those that stopped treatment, were lost to follow up, those that died during follow up from TB, those that died during follow up were TB contributed to their death, and those who were still on treatment. Those that were not evaluated were treated as missing.', 
                  'Assessed via follow up at 12 and 24 months: mortality is defined as cases with an overall outcome of death, and survival is defined as those that completed treatment, were still on treatment, and stopped treatment. Those that were lost to follow up, or not evaluated were treated as missing', 
                  'For cases with an overall outcome of died, and whose cause of death was known to be TB or to be related to TB. Those that were known to have not died, or who were known to have died from a cause other than from TB were defined to have not died from TB.',
                  'Death due to TB is defined as those that died directly from TB, or where TB had contributed to their death with death not due to TB being cases that died from any other cause. Conditioned on all-cause mortality, for cases with a known cause of death.')

@@

Missing data can take several forms, data that are missing completely at random (MCAR), data that are missing at random (MAR) and data that are missing not at random (MNAR).[@Sterne2009a] Data that are MAR are missing with a mechanism that is conditional on observed variables, whilst MNAR are missing with a mechanism that is conditional on variables that are not observed. Data that is MAR, and MNAR may lead to biases when analysing the data, however it is not possible to deduce from the observed data what the mechanism driving missing data is. Therefore, it is necessary to account for these potential biases during the analysis stage. This is possible using a variety of methods such as scenario analysis accounting for the 'best' and 'worst' case scenarios, and multiple imputation of missing data using additional variables in the dataset to inform the imputation model.[@Sterne2009a]

As the ETS is aggregated across England, from a variety of sources, some level of missing data are inevitable. This takes two forms: under-reporting of notified cases, of which there is some evidence in the literature,[@Pillaye2003] and data missing for a notified case. The former is particularly problematic as apart from using comparative studies the characteristics of those that are not notified is unknown. For variables that are missing data within the data-set it is possible to calculate the proportion of missing data (Figure \@ref(fig:plot-missing-struct), Table \@ref(tab:missing-var-tabs)) but care must be taken to account for nested variables such as date of death and year of BCG vaccination. This can be done by assuming that the nested variables takes the value of the top level variable when it is known that the variable is not truly missing. An example of this is using overall outcome for date of death when notifications are known to have not died.  Doing this shows high completeness for common demographic variables such as sex, age, ethnic group and UK birth status. More problematically, BCG status and year of BCG status have a high percentage missing, even after accounting for the introduction of national collection of these variables in 2008. Socio-economic status (as national quintiles) was not collected until 2010 but after this point is highly complete.  Comparing pre 2009 and post 2008 in Table \@ref(tab:missing-var-tabs) (and by inspecting Figure \@ref(fig:plot-missing-struct)) there are also issues of changing completeness over time,[@PHE2016; @PHE2017] if this is not accounted for than it may lead to spurious trends. Figure \@ref(fig:plot-missing-struct) also indicates that there are multiple groups of variables that share a common pattern of missing data.

```{r plot-missing-struct, fig.scap = "Summary plot of missing data in the extract of the ETS data used in this thesis.", fig.cap = "Summary plot of missing data in the extract of the ETS data used in this thesis. Due to the large size of the dataset, the data has been sub-sampled with only 20\\% of the data shown in this figure. Notifications have been ordered by date of notification from left to right. The following subset of variables are shown: year (year), sex (sex), age (age), PHE Centre (phec), Occupation (occat), Ethnic group (ethgrp), UK birth status (ukborn), Time since entry (timesinceent), date of symptom onset (symptonset), date of diagnosis (datediag), started treatment (startedtreat), date of starting treatment (starttreatdate), treatment end date (txenddate), pulmonary or extra-pulmonary TB (pulmextrapulm), culture (culture), sputum smear status (sputsmear), drug resistance (anyres), previous diagnosis (prevdiag), BCG status(bcgvacc), Year of BCG vaccination (bcgvaccyr), overall outcome (overalloutcome), cause of death (tomdeathrelate), socio-economic status quintiles (natquintile), and date of death (dateofdeath). Nested variables have been accounted for (i.e date of death has had an entry added for cases that are known to have not died), so that true missingness for all variables is estimated.", out.width = "60%", out.extra = ""}
knitr::include_graphics("chapters/tb-epi-england/figures/plot-missing-struct.png")
```

@@

kable(miss_ets,
      caption.short = "Breakdown of missing data from the ETS prior to the web based system (pre 2009) and post (post 2008) by variable, ordered by the  percentage missing for a subset of variables.",
      caption = "Breakdown of missing data from the ETS prior to the web based system (pre 2009) and post (post 2008) by variable, ordered by the  percentage missing for a subset of variables. The following subset of variables are shown year (year), sex (sex), age (age), PHE Centre (phec), Occupation (occat), Ethnic group (ethgrp), UK birth status (ukborn), Time since entry (timesinceent), date of symptom onset (symptonset), date of diagnosis (datediag), started treatment (startedtreat), date of starting treatment (starttreatdate), treatment end date (txenddate), pulmonary or extra-pulmonary TB (pulmextrapulm), culture (culture), sputum smear status (sputsmear), drug resistance (anyres), previous diagnosis (prevdiag), BCG status(bcgvacc), Year of BCG vaccination (bcgvaccyr), overall outcome (overalloutcome), cause of death (tomdeathrelate), socio-economic status quintiles (natquintile), and date of death (dateofdeath). Nested variables have been accounted for (i.e data of death has had an entry added for cases that are known to have not died), so that true missingness for all variables is estimated.",
      booktabs = TRUE) %>%
  add_header_above(c(" ", "Pre 2009" = 2, "Post 2008" = 2)) %>% 
  kable_styling(latex_options = "HOLD_position") 
```

For nested variables with rare outcomes assuming the top level variable value can mask the underlying amount of missing data. An alternative approach is to filter the data for the top level variable required for the nested variable to be defined and to then compute the proportion of these notifications that are missing data for the outcome of interest. For the date of starting treatment this approach leads to an estimate of `r missing_stats[["date_treat"]]` being missing, which is more complete than previously estimated. For cases that are known to have completed treatment `r missing_stats[["date_treat_end"]]` are missing a date for the end of treatment. In notifications that are known to have died, `r missing_stats[["date_death"]]` were missing the date of death and `r missing_stats[["cause_death"]]` were missing the cause of death. In any analysis were these variables are used the missing data for these variables will need to be carefully adjusted for. In particular, if cause of death is used it must be clearly stated that it is highly missing and results based on this variable should be properly caveated.

@@

As previously discussed, missing data may be MAR or MNAR, which may introduce biases into any analyses based on these data. This is of particular importance for variables that have high levels of missingness, as any introduced bias is likely to have a greater impact on the overall results, and for variables that are used extensively in analyses later in this thesis. Unfortunately MNAR data cannot be detected, so bias from this source cannot be discounted. However, it is possible to detect potential MAR mechanisms from observed variables that would not necessarily be included in a model used for analysis. Although these associations may themselves be caused by an external factor. In the following section I explore variables associated with data being missing for several key variables including: BCG status, year of BCG vaccination, date of death, cause of death, date of symptom onset, date of diagnosis, date of starting treatment and date of ending treatment. All of these variables were shown to have high levels of missing data in the previous section and they will all be used extensively throughout this thesis.

In order to explore the drivers of missing data I have reformulated the problem as a logistic regression for each variable of interest, with the outcome being data completeness (complete/missing). This allows variables that are hypothesised to be related to missing data to be adjusted for and their independent impact on data completeness to be estimated. Unlike classic approaches to missing data, such as multiple imputation by chained regression (MICE),[@Groothuis-oudshoorn] this is not an imputation. The details of the approach are discussed below.

@@

In order to reformulate missing data as a logistic regression I took the following steps:

@@

1. Fit a logistic regression model with the temporary data status variable as the outcome, adjusting for the hypothesised drivers of missingness.

1. Exponentiate the returned coefficients, and confidence intervals so that they represent Odds Ratios (ORs).

@@

For all outcomes considered I adjusted for the same set of demographic variables that were both highly complete and also plausibly linked to missingness for all outcomes considered. These were: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles). For socio-economic group 1 indicates the most deprived quintile. Complete case analysis has been used, with the dataset limited to notifications from 2010 and on-wards as socio-economic status was not collected prior to this.

@@

It is clear that BCG status is missing with a MAR mechanism for the variables considered (Table \@ref(tab:bcgvacc-miss)). BCG data missingness is strongly associated with year of notification, sex age, ethnic group, and socio-economic status. It appears that after adjusting for other variables data completeness increased from 2010 until 2012 but has since showed no clear trend. Men appear to be more likely than women to have a missing BCG status, with the non-UK born also being more likely than the UK born to be missing BCG status. The proportion of those missing BCG status increases with age, with those aged 65+ being over 4 times more likely to be missing BCG status than those aged 0-14 years old. There is also evidence to suggest that notifications in the lowest socio-economic group are more likely to have a missing BCG status but there was no clear evidence of a trend across socio-economic quintiles. The White ethnic group was more likely to have a missing BCG status than any other ethnic group.

```{r bcgvacc-miss}
kable(model_miss_results[["bcgvacc"]],
      caption.short = "Results from a logistic regression model with data completeness (Complete/Missing) for BCG vaccination as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles).",
      caption = "Results from a logistic regression model with data completeness (Complete/Missing) for BCG vaccination as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles). For socio-economic group 1 indicates the most deprived quintile. Notifications from 2010 onwards were included as socio-economic status was not collected before this. Complete case analysis weas used. Odds ratios shown are adjusted for all explanatory variables. The model indicates that BCG status is missing at random for the variables considered.", booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(font_size = 8, 
                latex_options = c("repeat_header")) %>% 
  column_spec(c(1), width = "1.5cm") %>% 
  column_spec(c(4), width = "2cm") %>% 
  column_spec(c(6, 7), width = "1.5cm") %>% 
  kable_styling(latex_options = "hold_position")
```

@@

```{r bcgvaccyr-miss}
kable(model_miss_results[["bvcvaccyr"]], format = "latex",
      caption.short = "Results from a logistic regression model with data completeness (Complete/Missing) for year of BCG vaccination as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles).",
      caption = "Results from a logistic regression model with data completeness (Complete/Missing) for year of BCG vaccination as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles). For socio-economic group 1 indicates the most deprived quintile. Notifications from 2010 onwards were included as socio-economic status was not collected before this. Complete case analysis weas used. Odds ratios shown are adjusted for all explanatory variables. The model indicates that year of BCG vaccination is missing at random for the variables considered.", booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(font_size = 8, latex_options = c("repeat_header")) %>% 
  column_spec(c(1), width = "1.5cm") %>% 
  column_spec(c(4), width = "2cm") %>% 
  column_spec(c(6, 7), width = "1.5cm")
```

@@

```{r dateofdeath-miss}
kable(model_miss_results[["dateofdeath"]],
      caption.short = "Results from a logistic regression model with data completeness (Complete/Missing) for date of death as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles).",
      caption = "Results from a logistic regression model with data completeness (Complete/Missing) for date of death as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles). For socio-economic group 1 indicates the most deprived quintile. Notifications from 2010 onwards were included as socio-economic status was not collected before this. Complete case analysis weas used. Odds ratios shown are adjusted for all explanatory variables. The model indicates that there is some evidence that date of death is missing at random for ethnic group, with weaker evidence for all other variables.", booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(font_size = 8, latex_options = c("repeat_header")) %>% 
  column_spec(c(1), width = "1.5cm") %>% 
  column_spec(c(4), width = "2cm") %>% 
  column_spec(c(6, 7), width = "1.5cm") %>% 
  kable_styling(latex_options = "hold_position")
```

@@

```{r tomdeathrealat-miss}
kable(model_miss_results[["tomdeathrealat"]],
      caption.short = "Results from a logistic regression model with data completeness (Complete/Missing) for cause of death as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles).",
      caption = "Results from a logistic regression model with data completeness (Complete/Missing) for cause of death as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles). For socio-economic group 1 indicates the most deprived quintile. Notifications from 2010 onwards were included as socio-economic status was not collected before this. Complete case analysis weas used. Odds ratios shown are adjusted for all explanatory variables. The model indicates that cause of death is missing at random for ethnic group and UK birth status, with little evidence for any other variables", booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(font_size = 8, latex_options = c("repeat_header")) %>% 
  column_spec(c(1), width = "1.5cm") %>% 
  column_spec(c(4), width = "2cm") %>% 
  column_spec(c(6, 7), width = "1.5cm") %>% 
  kable_styling(latex_options = "hold_position")
```
